---
title: "Lab 5: Bikeshare SpaceTime Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Parker Otto"
date: today
format:
  html:
    code-fold: false
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: darkly
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: **rebalancing bikes to meet anticipated demand**.

Imagine you're an Indego operations manager at 6:00 AM on a Monday morning. You have: - 200 stations across Philadelphia - Limited trucks and staff for moving bikes - 2-3 hours before morning rush hour demand peaks - **The question:** Which stations will run out of bikes by 8:30 AM?

This lab will teach you to build predictive models that forecast bike share demand across **space** (different stations) and **time** (different hours) to help solve this operational problem.

## Learning Objectives

By the end of this assignment, you will be able to:

1.  **Understand panel data structure** for space-time analysis
2.  **Create temporal lag variables** to capture demand persistence
3.  **Build multiple predictive models** with increasing complexity
4.  **Validate models temporally** (train on past, test on future)
5.  **Analyze prediction errors** in both space and time
6.  **Engineer new features** based on error patterns
7.  **Critically evaluate** when prediction errors matter most

## Assignment Structure

**Part 1 (In-Class/Together):** Work through Q1 2025 data following this code

**Part 2 (HW5 - Teams of 2):** - Download a different quarter of Indego data (Q2, Q3, or Q4 2025) - Apply the same workflow to your quarter - Analyze error patterns in your quarter - Add 2-3 new features to improve the model - Write a brief report on what you learned

#SET UP

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}

census_api_key(Sys.getenv("a05b265afea2fb4dfbd350edadaf50fb30255de8"))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

# IMPORT AND PREP DATA

## Load Indego Trip Data (Q1 2025)

**Note for HW5:** You'll download a different quarter from: https://www.rideindego.com/about/data/

```{r load_indego}
# Read Q1 2025 data
indego <- read_csv("indego-trips-2025-q1.csv")

# Quick look at the data
glimpse(indego)
```

**Pause here** Look at the data. What do you suspect the coordinate system is? What do you notice about the date and time structure?

## Examine the Data Structure

```{r explore_data}
# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure. Read up on the lubridate library for more information on the date/time data extraction.

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

**Question** Explain the interval60 feature to someone else at your table.

#EDA OF INDEGO DATA

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

**Question:** What patterns do you see? How does ridership change over time?

Notice a big spike in Februrary of 2025 (the 14th - not just Valentine's Day)? Was anybody in Philly last Feb? What closed everything and cancelled my PPA class? ... Hint: Fly Eagles Fly

Let's investigate...compare the 14th to all other Friday's in the data...what did you find? Keep this in mind for future Feature Engineering.

```{r}
fly_eagles_fly <- daily_trips %>% filter(date == "2025-02-14")

typical_boring_friday <- indego %>%
  filter(dotw == "Fri", date != "2025-02-14") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(fly_eagles_fly)
print(typical_boring_friday)

```

## Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

**Question:** When are the peak hours? How do weekends differ from weekdays?

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

We'll get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

**Question** These data are in WGS84. If you wanted to calculate a spatial distance-based variable, like a buffer, should you use the dataset as it?

## Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

## Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations..

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

## Get Weather Data

Weather significantly affects bike share demand! Let's get hourly weather for Philadelphia.

```{r get_weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

Who is glad it's Spring!?

```{r visualize-weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

## Why Temporal Lags?

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

**Note:** We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

Temporal lags: good Spatial lags: bad!

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# TEMPORAL TRAIN/TEST SPLIT

**CRITICAL:** We must train on PAST data and test on FUTURE data!

## Why Temporal Validation Matters

In real operations, at 6:00 AM on March 15, we need to predict demand for March 15-31. We have data from Jan 1 - March 14, but NOT from March 15-31 (it hasn't happened yet!).

**Wrong approach:** Train on weeks 10-13, test on weeks 1-9 (predicting past from future!)

**Correct approach:** Train on weeks 1-9, test on weeks 10-13 (predicting future from past)

```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

#BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

-   All weekdays have positive coefficients (0.029 to 0.052)
-   Tuesday has the highest weekday effect (+0.052)
-   Weekdays likely benefit from concentrated commuting patterns

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (-0.061 and -0.065)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation 0 (baseline) 0.000 trips/hour (midnight) 1 -0.018 slightly fewer than midnight ... 6 +0.151 morning activity starting 7 +0.276 morning rush building 8 +0.487 PEAK morning rush 9 +0.350 post-rush ... 17 +0.568 PEAK evening rush (5 PM!) 18 +0.389 evening declining ... 23 +0.034 late night minimal

Isn't this fun!

## Model 2: Add Temporal Lags

```{r model2}
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

**Question:** Did adding lags improve R²? Why or why not?

## Model 3: Add Demographics

```{r model3}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

**What do station fixed effects capture?** Baseline differences in demand across stations (some are just busier than others!).

## Model 5: Add Rush Hour Interaction

```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question:** Which features gave us the biggest improvement?

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

**Question:** Where is the model performing well? Where is it struggling?

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 1, #so you can see
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 1, #so you can see
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

**Question:** Do you see spatial clustering of errors? What neighborhoods have high errors?

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

**Critical Question:** Are prediction errors systematically higher in certain demographic groups? What are the equity implications?

# LAB 5: YOUR TURN!

For Homework 5, you'll work either all byyyy yourself or in teams of 2 to:

## Part 1: Replicate with Different Quarter (alternately, you could do a longer time-span by merging multiple quarters together. I'm not picky about that)

1.  **Download data** for Q2, Q3, or Q4 2025 from: https://www.rideindego.com/about/data/

2.  **Adapt this code** to work with your quarter:

    -   Update date ranges for weather data
    -   Check for any data structure changes
    -   Create the same 5 models
    -   Calculate MAE for each model
    
```{r load-indego-q2}

indegoq2 <- read_csv("indego-trips-2025-q2.csv")

glimpse(indegoq2)

```

```{r explore-q2-data}

# How many trips?
cat("Total trips in Q1 2025:", nrow(indegoq2), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indegoq2$start_time)), "to", 
    max(mdy_hm(indegoq2$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indegoq2$start_station)), "\n")

# Trip types
table(indegoq2$trip_route_category)

# Passholder types
table(indegoq2$passholder_type)

# Bike types
table(indegoq2$bike_type)

```
```{r create-q2-time-bins}
indegoq2 <- indegoq2 %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indegoq2 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

```{r q2-trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips_q2 <- indegoq2 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips_q2, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2025",
    subtitle = "Spring demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

```{r hourly_patterns_q2}
# Average trips by hour and day type
hourly_patterns_q2 <- indegoq2 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns_q2, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns (Q2)",
    subtitle = "Clear commute patterns on Q2 weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```
```{r top_stations-q2}
# Most popular origin stations
top_stations_q2 <- indegoq2 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations_q2, 
      caption = "Top 20 Indego Stations by Trip Origins (Q2)",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r q2-map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indegoq2,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```
```{r join_census_to_stations-q2}

# Create sf object for stations
stations_sf_q2 <- indegoq2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census_q2 <- st_join(stations_sf_q2, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map_q2 <- indegoq2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census_q2 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census_q2 <- indegoq2 %>%
  left_join(
    stations_census_q2 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

# Prepare data for visualization
stations_for_map_q2 <- indegoq2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census_q2 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map_q2 %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map_q2 %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Q2 Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme

```
```{r keep-stations-q2}
# Identify which stations to keep
valid_stations_q2 <- stations_census_q2 %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census_q2 <- indegoq2 %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census_q2 %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

```{r get_weather-q2}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data_q2 <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-04-01",
  date_end = "2025-06-30"
)

# Process weather data
weather_processed_q2 <- weather_data_q2 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete_q2 <- weather_processed_q2 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete_q2 %>% select(Temperature, Precipitation, Wind_Speed))
```

```{r visualize-weather-q2}
#| message: false
#| warning: false
#| 
ggplot(weather_complete_q2, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2025",
    subtitle = "Early to late spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

```{r aggregate-trips-q2}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel_q2 <- indego_census_q2 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel_q2)

# How many unique stations?
length(unique(trips_panel_q2$start_station))

# How many unique hours?
length(unique(trips_panel_q2$interval60))

hist(trips_panel_q2$Trip_Count)
```
```{r complete-panel-q2}
# Calculate expected panel size
n_stations_q2 <- length(unique(trips_panel_q2$start_station))
n_hours_q2 <- length(unique(trips_panel_q2$interval60))
expected_rows_q2 <- n_stations_q2 * n_hours_q2

cat("Expected panel rows:", format(expected_rows_q2, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel_q2), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows_q2 - nrow(trips_panel_q2), big.mark = ","), "\n")

# Create complete panel
study_panel_q2 <- expand_grid(
  interval60 = unique(trips_panel_q2$interval60),
  start_station = unique(trips_panel_q2$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel_q2 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel_q2)
hist(study_panel_q2$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes_q2 <- trips_panel_q2 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes_q2)

study_panel_q2 <- study_panel_q2 %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel_q2)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel_q2), big.mark = ","), "\n")
```

```{r add_time_features-q2}
study_panel_q2 <- study_panel_q2 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel_q2)
```

```{r join_weather-q2}
study_panel_q2 <- study_panel_q2 %>%
  left_join(weather_complete_q2, by = "interval60")

glimpse(study_panel_q2)

# Check for missing values
summary(study_panel_q2 %>% select(Trip_Count, Temperature, Precipitation))
```

```{r create_lags-q2}
# Sort by station and time
study_panel_q2 <- study_panel_q2 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel_q2 <- study_panel_q2 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete_q2 <- study_panel_q2 %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

```{r lag_correlations-q2}
# Sample one station to visualize
example_station_q2 <- study_panel_complete_q2 %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station_q2, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station (Q2)",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

```{r temporal-split-q2}
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Q2 has weeks 14-26 (Apr-Jun)
# Train on weeks 14-22
# Test on weeks 23-26

# Which stations have trips in BOTH early and late periods?
early_stations_q2 <- study_panel_complete_q2 %>%
  filter(week < 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations_q2 <- study_panel_complete_q2 %>%
  filter(week >= 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations_q2 <- intersect(early_stations_q2, late_stations_q2)

# Filter panel to only common stations
study_panel_complete_q2 <- study_panel_complete_q2 %>%
  filter(start_station %in% common_stations_q2)

# NOW create train/test split
train_q2 <- study_panel_complete_q2 %>%
  filter(week < 23)

test_q2 <- study_panel_complete_q2 %>%
  filter(week >= 23)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

```{r model1-q2}

# Create day of week factor with treatment (dummy) coding
train_q2 <- train_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train_q2$dotw_simple) <- contr.treatment(7)

# Now run the model
model1_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train_q2
)

summary(model1_q2)
```

```{r model2-q2}
model2_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train_q2
)

summary(model2_q2)
```

```{r model3-q2}
model3_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train_q2
)

summary(model3_q2)
```
```{r model4-q2}
model4_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train_q2
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4Q2 R-squared:", summary(model4_q2)$r.squared, "\n")
cat("Model 4Q2 Adj R-squared:", summary(model4_q2)$adj.r.squared, "\n")
```
```{r model5-q2}
model5_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train_q2
)

cat("Model 5Q2 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5Q2 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

```{r calculate_mae-q2}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test_q2 <- test_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test_q2$dotw_simple) <- contr.treatment(7)

test_q2 <- test_q2 %>%
  mutate(
    pred1 = predict(model1_q2, newdata = test_q2),
    pred2 = predict(model2_q2, newdata = test_q2),
    pred3 = predict(model3_q2, newdata = test_q2),
    pred4 = predict(model4_q2, newdata = test_q2),
    pred5 = predict(model5_q2, newdata = test_q2)
  )

# Calculate MAE for each model
mae_results_q2 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test_q2$Trip_Count - test_q2$pred1), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred3), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred4), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred5), na.rm = TRUE)
  )
)

kable(mae_results_q2, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set) (Q2)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

3.  **Compare results** to Q1 2025:

    -   How do MAE values compare? Why might they differ?
    -   Are temporal patterns different (e.g., summer vs. winter)?
    -   Which features are most important in your quarter?

```{r q1-q2-comparison}

mae_results$Quarter <- "Q1"
mae_results_q2$Quarter <- "Q2"

mae_results_combined <- rbind(mae_results, mae_results_q2)

kable(mae_results_combined,
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)", "Quarter")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r temporal-q1q2}

dt_q1 <- ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Daily Ridership - Q1 2025",
    subtitle = "Spring demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  ylim(500, 6000) +
  plotTheme

dt_q2 <- ggplot(daily_trips_q2, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Daily Ridership - Q2 2025",
    subtitle = "Spring demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  ylim(500, 6000) +
  plotTheme

hp_q1 <- ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Avg Hourly Ridership (Q1)",
    subtitle = "Clear commute patterns on Q1 weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  ylim(0, 500) +
  plotTheme

hp_q2 <- ggplot(hourly_patterns_q2, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Avg Hourly Ridership (Q2)",
    subtitle = "Clear commute patterns on Q2 weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  ylim(0, 500) +
  plotTheme

(dt_q1 | dt_q2) / (hp_q1 | hp_q2)

```

The pattern of MAE with the addition of new data between Q2 (spring) and Q1 (winter) trained models generally follows a similar trend, but the Q2 data produced models with higher overall MAE (i.e. models that performed worse) than the Q1 data models.  Additionally, including rush hour interaction with Q2 models actually led to a worse MAE than with Q1 (perhaps this could have something to do with rush hour participation differences between winter, when there is school, and late spring, when there may not be school for some). The temporal change in ridership during the weekday also changed between Q1 and Q2.  While the peak AM period saw a ~100 trip increase, the PM rush period saw a ~200 trip increase.  This could potentially be due to workers being more likely to take a leisurely bike home after sitting in the office all day (as opposed to biking in the morning, when they're potentially tired or wary of getting sweaty).  In Q2, it would appear that station fixed effects are the most important feature, as they increase adjusted R-squared without impacting MAE.

## Part 2: Error Analysis

Analyze your model's errors in detail:

1.  **Spatial patterns:**

    -   Create error maps
    -   Identify neighborhoods with high errors
    -   Hypothesize why (missing features? different demand patterns?)

```{r obs_vs_pred-q2}
#| message: false
#| warning: false

test_q2 <- test_q2 %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test_q2, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips (Q2)",
    subtitle = "Model 2Q2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

```{r spatial_errors-q2}
# Calculate MAE by station
station_errors_q2 <- test_q2 %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors_q2 <- test_q2 %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1_q2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors_q2,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 1, #for better visibility
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors Q2") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2_q2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors_q2,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 1, #for better visibility
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand Q2") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1_q2 | p2_q2)

```

Prediction errors in Q2 2025 appear to be clustered in Center City.  Given that this clustering pattern matches a map of average demand for the same quarter, one hypothesis for why errors are clustered in Center City is that the model struggles to predict areas with high demand.  

2.  **Temporal patterns:**

    -   When are errors highest?
    -   Do certain hours/days have systematic under/over-prediction?
    -   Are there seasonal patterns?



```{r temporal_errors-q2}
# MAE by time of day and day type
temporal_errors_q2 <- test_q2 %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors_q2, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period Q2",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Temporally, there appear to be more consistent errors during the evening and mid-day with more errors during the PM rush and weekday AM rush.  Overnight shows the lowest error amount with both weekend and weekday errors fairly consistent.  This pattern of errors somewhat matches demand (in that low errors and low demand are true for overnight and high error high demand is true for PM rush, evening, and AM weekday rush).  However, this demand pattern does not account for fairly high mid-day errors, as there is generally a cratering in the amount of rides taken at this time.  

3.  **Demographic patterns:**

    -   Relate errors to census characteristics
    -   Are certain communities systematically harder to predict?
    -   What are the equity implications?
    
```{r errors_demographics-q2}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo_q2 <- station_errors_q2 %>%
  left_join(
    station_attributes_q2 %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
ed_p1_q2 <- ggplot(station_errors_demo_q2, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

ed_p2_q2 <- ggplot(station_errors_demo_q2, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

ed_p3_q2 <- ggplot(station_errors_demo_q2, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(ed_p1_q2 | ed_p2_q2) / (ed_p3_q2 | plot_spacer())
```

According to these graphs, the model has trouble predicting for areas in which the following groups live: high median income, low transit commuters, and majority white neighborhoods.  Interestingly, this appears to be the opposite problem that many of the other models built during this class face.  Perhaps this could be due to bike usage being more unpredictable for white, high-earning areas with low transit usage given the nature of those spaces.  These would likely be suburban and car-centric which would lead to people either not really biking (as it could be dangerous to do so) and if they do bike, it could be due to the presence of factors not included in the model (like rail-to-trails or dedicated bike lanes).

## Part 3: Feature Engineering & model improvement

Based on your error analysis, add 2-3 NEW features to improve the model:

**Potential features to consider:**

*Temporal features:*

-   Holiday indicators (Memorial Day, July 4th, Labor Day)
-   School calendar (Penn, Drexel, Temple in session?)
-   Special events (concerts, sports games, conventions)
-   Day of month (payday effects?)

*Weather features:*

-   Feels-like temperature (wind chill/heat index)
-   "Perfect biking weather" indicator (60-75°F, no rain)
-   Precipitation forecast (not just current)
-   Weekend + nice weather interaction

*Spatial features:*

-   Distance to Center City
-   Distance to nearest university
-   Distance to nearest park
-   Points of interest nearby (restaurants, offices, bars)
-   Station capacity
-   Bike lane connectivity

*Trip history features:*

-   Rolling 7-day average demand
-   Same hour last week
-   Station "type" clustering (residential, commercial, tourist)

**Implementation:**

-   Add your features to the best model
-   Compare MAE before and after
-   Explain *why* you chose these features
-   Did they improve predictions? Where?

**Try a poisson model for count data**

-   Does this improve model fit?

```{r buffered-bike-lanes}

bike_net <- st_read("Bike_Network.geojson")

stations_sf_q2_reproj <- st_transform(stations_sf_q2, crs = 'EPSG:2272')

stations_buff_q2 <- st_buffer(stations_sf_q2_reproj, 2640) #1/2 mile buffer

bike_net_reproj <- st_transform(bike_net, 'EPSG:2272')
stations_buff_q2 <- st_transform(stations_buff_q2, 'EPSG:2272')  # make sure buffer is same CRS

stations_sf_q2_reproj <- stations_sf_q2_reproj %>%
  mutate(
    bl_len_halfmi = map_dbl(st_geometry(stations_buff_q2), function(buff) {
      buff_sfc <- st_sfc(buff, crs = 2272)
      st_intersection(bike_net_reproj, buff_sfc) %>%
        st_length() %>%
        sum()
    })
  )

#convert feet to mile

stations_sf_q2_reproj <- stations_sf_q2_reproj %>% 
  mutate(nearby_bl_len_mi = bl_len_halfmi / 5280)

```

``` {r dist-to-trails}

trail_net <- st_read("Existing_Trails.geojson")

trail_net <- st_transform(trail_net, 'EPSG:2272')

stations_sf_q2_reproj <- stations_sf_q2_reproj %>% 
  mutate(trail_dist = st_distance(geometry, st_union(trail_net)) %>% as.numeric() / 5280)

```

```{r}

study_panel_complete_q2 <- study_panel_complete_q2 %>% 
  left_join(st_drop_geometry(stations_sf_q2_reproj), by = "start_station")

train_q2 <- study_panel_complete_q2 %>%
  filter(week < 23)

test_q2 <- study_panel_complete_q2 %>%
  filter(week >= 23)

train_q2 <- train_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

```

```{r}

#append to model 4 since model 5 had worse mae

model6_q2 <- lm(
    Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    nearby_bl_len_mi + trail_dist + #2 new vars
    as.factor(start_station),
  data = train_q2
)

# append to model 5 shows better results

model7_q2 <- lm(
    Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    nearby_bl_len_mi + trail_dist + #2 new vars
    as.factor(start_station) + rush_hour * weekend,
  data = train_q2
)

# Coefficients without station dummies
library(broom)

cat("=== Model 7 Q2 - Key Predictors ===\n")
tidy(model7_q2) %>%
  filter(!grepl("start_station", term)) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  print(n = Inf)

# Model fit metrics
cat("\nR-squared:    ", summary(model7_q2)$r.squared, "\n")
cat("Adj R-squared:", summary(model7_q2)$adj.r.squared, "\n")
cat("Observations: ", nobs(model7_q2), "\n")

```
```{r calculate_mae-q2-with-m6}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test_q2 <- test_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test_q2$dotw_simple) <- contr.treatment(7)

test_q2 <- test_q2 %>%
  mutate(
    pred1 = predict(model1_q2, newdata = test_q2),
    pred2 = predict(model2_q2, newdata = test_q2),
    pred3 = predict(model3_q2, newdata = test_q2),
    pred4 = predict(model4_q2, newdata = test_q2),
    pred5 = predict(model5_q2, newdata = test_q2),
    pred6 = predict(model6_q2, newdata = test_q2),
    pred7 = predict(model7_q2, newdata = test_q2)
  )

# Calculate MAE for each model
mae_results_q2_wm6 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction",
    "6. + Bike Lane Len. Within Half Mi. + Trail Dist. - M5 Addn.",
    "7. M6 + M5 Addn."
  ),
  MAE = c(
    mean(abs(test_q2$Trip_Count - test_q2$pred1), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred3), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred4), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred5), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred6), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred7), na.rm = TRUE)
  )
)

kable(mae_results_q2_wm6, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set) (Q2)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Part 4: Critical Reflection

Write 1-2 paragraphs addressing:

1.  **Operational implications:**
    -   Is your final MAE "good enough" for Indego to use?
    -   When do prediction errors cause problems for rebalancing?
    -   Would you recommend deploying this system? Under what conditions?
2.  **Equity considerations:**
    -   Do prediction errors disproportionately affect certain neighborhoods?
    -   Could this system worsen existing disparities in bike access?
    -   What safeguards would you recommend?
3.  **Model limitations:**
    -   What patterns is your model missing?
    -   What assumptions might not hold in real deployment?
    -   How would you improve this with more time/data?

After building these several models to predict station-level Indego trips, I would say that the final model's MAE of 0.69 is a low enough number to be a "good enough" model Indego to use (predicting on average +/- 0.69 trips is not a huge number, all else considered).  What I am not saying, though, is that this model *should* be used to place bike stations.  While environmental and temporal factors do show to impact the count of trips, philosophically these must not be the deciding factors on where to place access to bikes (after all, an Indego station during evening rush located in the middle lane of I-95 would not get any use).  While I attempted to capture more spatial variables with the two additions for models 6 and 7 (bike lane length within half a mile of the station, as in theory more people would ride a bike if they were on/near a bike lane, and trail distance, as in theory more people would ride a bike if they were near a trail), they are not even close to an exhaustive list of what makes biking attractive to individuals.  Further models should include data from OnTheMap to get average distance for work commutes per census tract that each Indego station is in, car crash data (either the count of crashes within a buffer or count of specifically crashes involving bikes and pedestrians).  Additional data that could be included are, based on my own experience with bikeshare, average distance from end station (or availability of docking stations within a buffer) as the biggest factor for me personally on if I take bikeshare is whether my destination has a dock close by.  Further, this model's training data is far from complete, given that it only accounts for one quarter at a time.  Perhaps it would be beneficial to plan for peak usage (spring-fall) and simply omit winter months (Dec-Feb) under the assumption that no matter how good the bike network, stations will be un(der)used during these times (although notably that is not the case in some European contexts).  I chose Q2 data as it was simply the next on the list, but further comparison of different segments of this data could be interesting (e.g. models trained on just winter, just march, the whole year, or Mar-Nov).

At the end of the day, the model performs poorly in the high-demand, high-trip areas so there is relatively little risk of worsening existing disparities in bike access for disinvested neighborhoods and communities.  It is important for city leaders, planners, and the owners of Indego to understand that plopping a bike share station anywhere cannot make up for (1) lasting stigmas against cyclists, (2) valid fears of harm or death from car crashes or road rage, (3) car-centric scale and design, and much, much more.   To address the argument in favor of expanding Indego at all cost (e.g. "build it and they will come"), it is important to recognize that one would not build a series of dispersed parking lots complete with rental cars without building a high-quality network of streets to connect them.  Therefore, it is irresponsible to call for the simple expansion of bikeshare without first addressing the key barrier to Philadelphians getting on bikes in the first place: safety.

