---
title: "week6"
format: html
---

```{r}

# Load packages and data
library(tidyverse)
library(sf)
library(here)

# Load Boston housing data
boston <- read_csv(here("data/boston.csv"))

# Quick look at the data
glimpse(boston)


```


Simple Model:

```{r}

# Simple model: Predict price from living area
baseline_model <- lm(SalePrice ~ LivingArea, data = boston)

summary(baseline_model)

```

```{r}

# Add number of bathrooms
better_model <- lm(SalePrice ~ LivingArea + R_FULL_BTH, data = boston)
summary(better_model)

# Compare models
cat("Baseline R²:", summary(baseline_model)$r.squared, "\n")

cat("With bathrooms R²:", summary(better_model)$r.squared, "\n")

```

```{r}

library(sf)

# Convert boston data to sf object -- already has lat/long fields, so this is possible
# Transform to local coordinate system
boston.sf <- boston %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(2249)  # MA State Plane (feet)

# Load neighborhood boundaries
nhoods <- read_sf(here("data/BPDA_Neighborhood_Boundaries.geojson")) %>%
  st_transform(2249)  # Match CRS!

#nrow(nhoods)  # How many neighborhoods?

# Spatial join: Assign each house to its neighborhood
boston.sf <- boston.sf %>%
  st_join(nhoods, join = st_intersects)

# Check results
boston.sf %>%
  st_drop_geometry() %>%
  count(name) %>%
  arrange(desc(n))

```




```{r}

# Map neighborhoods with median prices
price_by_nhood <- boston.sf %>%
  st_drop_geometry() %>%
  group_by(name) %>%
  summarize(
    median_price = median(SalePrice, na.rm = TRUE),
    n_sales = n()
  )

# Join back to spatial data
nhoods_prices <- nhoods %>%
  left_join(price_by_nhood, by = "name")

# Create custom price classes
nhoods_prices <- nhoods_prices %>%
  mutate(
    price_class = cut(median_price,
                     breaks = c(0, 400000, 600000, 800000, 1000000, Inf),
                     labels = c("Under $400k", "$400k-$600k", "$600k-$800k", 
                               "$800k-$1M", "Over $1M"),
                     include.lowest = TRUE)
  )


# YlOrRd (Yellow-Orange-Red) - classic graduated
ggplot() +
  geom_sf(data = nhoods_prices, aes(fill = price_class), 
          color = "white", size = 0.5) +
  scale_fill_brewer(
    name = "Median Price",
    palette = "YlOrRd",  # Try also: "Reds", "OrRd", "YlGnBu", "PuRd"
    na.value = "grey90",
    direction = 1  # Use -1 to reverse (dark = low)
  ) +
  labs(
    title = "Median Home Prices by Boston Neighborhood",
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold")
  )





```

```{r}
# Which neighborhoods are most expensive?
price_by_nhood %>%
  arrange(desc(median_price)) %>%
  head(5)

```


```{r}
# Which have most sales?
price_by_nhood %>%
  arrange(desc(n_sales)) %>%
  head(5)
```

```{r}

# Ensure name is a factor (i.e., categorical variable)
boston.sf <- boston.sf %>%
  mutate(name = as.factor(name))

# Check which is reference (first alphabetically)
#levels(boston.sf$name)[1]

model_neighborhoods <- lm(SalePrice ~ LivingArea + name, 
                          data = boston.sf)

# Show just first 10 coefficients
summary(model_neighborhoods)$coef[1:10, ]

```

Luxury Premium Hypothesis (Interaction Variables With Square Footage and Neighborhood Wealth)

Hypotheses:

Wealthy Neighborhood = $300+ per square foot

Working Class Neighborhood = $100-150 per square foot

```{r}

# Define wealthy neighborhoods based on median prices
wealthy_hoods <- c("Back Bay", "Beacon Hill", "South End", "Bay Village")

# Create binary indicator (factor)
boston.sf <- boston.sf %>%
  mutate(
    wealthy_neighborhood = ifelse(name %in% wealthy_hoods, "Wealthy", "Not Wealthy"),
    wealthy_neighborhood = as.factor(wealthy_neighborhood)
  )

# Check the split
boston.sf %>%
  st_drop_geometry() %>%
  count(wealthy_neighborhood)

```
MODEL 1: No Interaction (Parallel Slope)

* Assumes that living area has the same effect in all neighborhoods
* Only the intercept differs (wealthy areas start higher)
* Parallel lines on a plot

```{r}

# Model assumes same slope everywhere
model_no_interact <- lm(SalePrice ~ LivingArea + wealthy_neighborhood, 
                        data = boston.sf)

summary(model_no_interact)$coef

```

MODEL 2: With Interaction (Different Slopes)

* Allows living area to have different effects in different neighborhoods.
* Both intercept AND slope differ.
* Non-parallel lines on a plot.

```{r}

# Model allows different slopes
model_interact <- lm(SalePrice ~ LivingArea * wealthy_neighborhood, 
                     data = boston.sf)

summary(model_interact)$coef

```
Interpreting the Interaction Coefficients:

Term	Coefficient	t-value	p-value	Sig
Intercept (Not Wealthy)	$358,542	19.235	0	***
Living Area (Not Wealthy)	$96	12.550	0	***
Wealthy Neighborhood Premium	-$377,937	-3.758	0	***
Extra \$/sq ft in Wealthy Areas	\$985	33.103	0	***

We get the un-intuitive negative premium here because that is an intercept adjustment (applies at 0 sqft). The slope difference (+985sq/ft) is huge - we can calculate when wealthy areas become more expensive (at what sq ft) = 384.

Breaking Down the Coefficients:

Not wealthy neighborhoods equation:
Price = \$358,542 + (\$96 * Living Area)
This means that for not wealthy neighborhoods, base price is \$358,542 and each additional sq ft adds \$96.

Wealthy Neighborhoods equation:
Price = -\$19,395 + (\$1,081 * Living Area)
Base price: -\$19,395 Each sq ft adds: \$1,081

Key takeaway: The Interaction Effect Wealthy areas value each sq ft $985 more than non-wealthy areas!

```{r}

#Visualizing the Interaction Effect:

#Notice that the lines in the graph below are NOT parallel (that's the interaction)

ggplot(boston.sf, aes(x = LivingArea, y = SalePrice, 
                      color = wealthy_neighborhood)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5) +
  scale_color_manual(
    values = c("Not Wealthy" = "#3498db", "Wealthy" = "#e74c3c"),
    name = "Neighborhood Type"
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1/1000000, suffix = "M")) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Living Area Effect Varies by Neighborhood Wealth",
    subtitle = "Different slopes = interaction effect (steeper line in wealthy areas)",
    x = "Living Area (sq ft)",
    y = "Sale Price",
    caption = "Notice: The red line (wealthy) is steeper than the blue line (not wealthy)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )

```


Compare Model Performance:

```{r}

# Compare R-squared

#Separate:

#cat("Model WITHOUT interaction R²:", round(summary(model_no_interact)$r.squared, 4), "\n")

#cat("Model WITH interaction R²:", round(summary(model_interact)$r.squared, 4), "\n")

#cat("Improvement:", round(summary(model_interact)$r.squared - summary(model_no_interact)$r.squared, 4), "\n")

#Combined:

cat(
  "Model WITHOUT interaction R²:", round(summary(model_no_interact)$r.squared, 4),
  "\nModel WITH interaction R²:", round(summary(model_interact)$r.squared, 4),
  "\nImprovement:", round(summary(model_interact)$r.squared - summary(model_no_interact)$r.squared, 4)
)

```

Interpretation of Improvement: We explain 16.35% more variation in prices by allowing different slopes!

Policy Implications:

* Boston operates as two distinct housing markets
  * Luxury Market: Every additional sq ft adds \$1,081
  * Standard Market: Space is valued, but way lower premium (\$96/sq ft)
  
* Affordability Crisis: The Interaction Amplifies Inequality
  * Large homes in wealthy areas become exponentially more expensive
  * Creates barriers to mobility between neighborhoods

* Policy Design: One-size-fits-all Policies May Fail
  * Property tax assessments should account for neighborhood-specific valuation
  * Housing assistance needs vary dramatically by area
  
When NOT To Use Interactions:

* Small samples
  * Need sufficient data in each group
    *Rough guideline is to try for at least 30 in each group, but preference is for more.

* Overfitting
  *Too many interactions make models unstable
  
POLYNOMIAL TERMS: Non-Linear Relationships: When straight lines don't fit

Signs of Non-Linearity:
* Curved residual plots
* Diminishing returns
* Accelerating effects
* U-shaped or inverted U patterns
  * Use log for monotonic patterns (i.e. always increasing / decreasing)
* Examples:
  * House age (depreciation and then vintage premium)
  * Test scores (plateau after studying)
  * Advertising (diminishing returns)
  * Crime prevention (early gains, then plateaus)

THEORY: The U-Shaped Age Effect

* New builds are expensive, medium-old builds are dated / not worth it, historical homes are vintage and rare, thus expensive.


```{r}

#Explore Age Effect U-Shape

# Calculate age from year built
boston.sf <- boston.sf %>%
  mutate(Age = 2025 - YR_BUILT)%>% filter(Age <2000)


# Check the distribution of age
#summary(boston.sf$Age)

# Visualize age distribution
ggplot(boston.sf, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of House Age in Boston",
       x = "Age (years)",
       y = "Count") +
  theme_minimal()

```

```{r}

# Simple linear relationship
model_age_linear <- lm(SalePrice ~ Age + LivingArea, data = boston.sf)

summary(model_age_linear)$coef

```

Interpretation: Each additional year of age changes price by $2834.01 (assumed constant rate)

```{r}
#| eval: true
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Plot with linear fit
ggplot(boston.sf, aes(x = Age, y = SalePrice)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1/1000, suffix = "k")) +
  labs(
    title = "House Age vs. Price: Linear (red) vs. Flexible Curve (blue)",
    x = "Age (years)",
    y = "Sale Price",
    caption = "Red = linear assumption, Blue = data-driven curve"
  ) +
  theme_minimal()

```

Adding A Polynomial Term: Age Squared

```{r}

# Quadratic model (Age²)
model_age_quad <- lm(SalePrice ~ Age + I(Age^2) + LivingArea, data = boston.sf)

#Note the use of I() which tells R to square Age (I means "as-is").
# lm(SalePrice ~ Age^2) # R reads this as just Age - Age interacted with itself up to degree 2, so Age^1, not Age squared 
# lm(SalePrice ~ I(Age^2)) # R reads this as Age squared

summary(model_age_quad)$coef

```

Interpretation: Model equation: Price = \$570,397 + (-\$13,008 * Age) + (\$80 * Age²) + (\$204 * LivingArea)

* ⚠️ Can’t Interpret Coefficients Directly!

With Age², the effect of age is no longer constant. You need to calculate the marginal effect. Marginal effect of Age = β₁ + 2×β₂×Age This means the effect changes at every age!

Compare model improvement:

```{r}

# R-squared comparison
r2_linear <- summary(model_age_linear)$r.squared
r2_quad <- summary(model_age_quad)$r.squared

cat(
  "Linear model R²:", round(r2_linear, 4),
  "\nQuadratic model R²:", round(r2_quad, 4),
  "\nImprovement:", round(r2_quad - r2_linear, 4),
  "\n\n"
)

```
```{r}

# F-test: Is the Age² term significant?
# Quadratic should have a lower residual sum of squares, if not, did nothing
anova(model_age_linear, model_age_quad)

```

Check residual plot:

```{r}

# Compare residual plots
par(mfrow = c(1, 2))

# Linear model residuals
plot(
  fitted(model_age_linear),
  residuals(model_age_linear),
  main = "Linear Model Residuals",
  xlab = "Fitted Values", ylab = "Residuals",
  abline(h = 0, col = "red", lty = 2)
)

# Quadratic model residuals  
plot(
  fitted(model_age_quad),
  residuals(model_age_quad),
  main = "Quadratic Model Residuals",
  xlab = "Fitted Values", ylab = "Residuals",
  abline(h = 0, col = "red", lty = 2)
)

```

```{r}



```

PART 3: Creating Spatial Features

Tobler's First Law of Geography:
* “Everything is related to everything else, but near things are more related than distant things”

What This Means for House Prices
* Crime nearby matters more than crime across the city
* Parks within walking distance affect value
* Your immediate neighborhood defines your market

How do we quantify “nearbyness” in a way our regression model can use?
* Create spatial features that measure proximity to amenities/disamenities

Three approaches to spatial features:
1. Buffer Aggregation
  * Count or sum events within a defined distance

2. k-Nearest Neighbors (kNN)
  * Average distance to k closest events

3. Distance to Specific Points
  * Straight-line distance to important locations

Explore these concepts with Boston crime data!

```{r}

# load data
bostonCrimes <- read_csv(here("data/bostonCrimes.csv"))
glimpse(bostonCrimes)
#previously set coordinate system, but if not use this code:
#boston.sf <- boston %>% 
#  st_as_sf(coords = c("Longitude", "Latitude"), crs = "EPSG:4326") %>%
#  st_transform(2249)

crimes.sf <- bostonCrimes %>%
  filter(UCR_PART == "Part One", Lat > -1, !is.na(Lat), !is.na(Long)) %>%
  st_as_sf(coords = c("Long", "Lat"), crs = 4326) %>%
  st_transform(2249) %>%
  distinct()

nhoods <- st_read(here("data/BPDA_Neighborhood_Boundaries.geojson")) %>%
  st_transform(2249)

# Quick check on projections - these should all be similar numbers
st_bbox(boston.sf)
st_bbox(crimes.sf)
st_bbox(nhoods)


# Check for missing coordinates
cat("Records with missing coordinates:", 
    sum(is.na(bostonCrimes$Lat) | is.na(bostonCrimes$Long)), "\n")

# Filter to serious and violent crimes (most relevant for housing prices)
violent_crimes <- bostonCrimes %>%
  filter(
    UCR_PART == "Part One",  # Serious crimes
    !is.na(Lat), !is.na(Long)  # Must have coordinates
  ) %>%
  select(OFFENSE_CODE_GROUP, OFFENSE_DESCRIPTION, SHOOTING, 
         Lat, Long, YEAR, OCCURRED_ON_DATE)

cat("Violent crime records:", nrow(violent_crimes), "\n")

# What types of crimes?
violent_crimes %>%
  count(OFFENSE_CODE_GROUP, sort = TRUE) %>%
  head(10)

```

Visualize Crime Locations:

```{r}
#| fig-width: 10
#| fig-height: 6

nhoods <- st_transform(nhoods, 2249)
# Plot to verify alignment
ggplot() +
  geom_sf(data = nhoods, fill = "white", color = "gray50", size = 0.8) +
  geom_sf(data = crimes.sf, alpha = 0.2, size = 0.3, color = "red") +
  geom_sf(data = boston.sf, alpha = 0.3, size = 0.5, color = "blue") +
  labs(
    title = "Checking Spatial Alignment",
    subtitle = "All layers should overlap - if not, CRS mismatch!",
    caption = "Red = crimes, Blue = house sales, Gray = neighborhoods"
  ) +
  theme_void()

```

Approach 1: Buffer Aggregation

```{r}

# Create buffer features - these will work now that CRS is correct
boston.sf <- boston.sf %>%
  mutate(
    crimes.Buffer = lengths(st_intersects(
      st_buffer(geometry, 660), #roughly 1/8 mile
      crimes.sf
    )),
    crimes_500ft = lengths(st_intersects(
      st_buffer(geometry, 500),
      crimes.sf
    ))
  )

# Check it worked
summary(boston.sf$crimes.Buffer)

# Quick histogram to verify
ggplot(boston.sf, aes(x = crimes.Buffer)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.7) +
  labs(title = "Crimes within 660ft of each house") +
  theme_minimal()

```

Approach 2: kNN 

```{r}

# Calculate distance matrix (houses to crimes)
dist_matrix <- st_distance(boston.sf, crimes.sf)

# Function to get mean distance to k nearest neighbors
get_knn_distance <- function(dist_matrix, k) {
  apply(dist_matrix, 1, function(distances) {
    # Sort and take first k, then average
    mean(as.numeric(sort(distances)[1:k]))
  })
}

# Create multiple kNN features
boston.sf <- boston.sf %>%
  mutate(
    crime_nn1 = get_knn_distance(dist_matrix, k = 1),
    crime_nn3 = get_knn_distance(dist_matrix, k = 3),
    crime_nn5 = get_knn_distance(dist_matrix, k = 5)
  )

# Check results
summary(boston.sf %>% st_drop_geometry() %>% select(starts_with("crime_nn")))

```

Interpretation: crime_nn3 = 273.29 means the average distance to the 3 nearest crimes is 273.29 feet.

Which k value correlates most with price?

```{r}

# Which k value correlates most with price?
# Drop geometry because can't be included in a correlation matrix
boston.sf %>%
  st_drop_geometry() %>%
  select(SalePrice, crime_nn1, crime_nn3, crime_nn5) %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  select(SalePrice)

```
Finding: The kNN feature with the strongest correlation tells us the relevant “zone of influence” for crime perception!

Approach 3: Distance to Downtown

```{r}

# Define downtown Boston (Boston Common: 42.3551° N, 71.0656° W)
downtown <- st_sfc(st_point(c(-71.0656, 42.3551)), crs = "EPSG:4326") %>%
  st_transform(2249)

# Calculate distance from each house to downtown
boston.sf <- boston.sf %>%
  mutate(
    dist_downtown_ft = as.numeric(st_distance(geometry, downtown)),
    dist_downtown_mi = dist_downtown_ft / 5280
  )

# Summary
summary(boston.sf$dist_downtown_mi)

```

All spatial features together summary:

```{r}

# Summary of all spatial features created
spatial_summary <- boston.sf %>%
  st_drop_geometry() %>%
  select(crimes.Buffer, crimes_500ft, crime_nn3, dist_downtown_mi) %>%
  summary()

spatial_summary

```
Interpretation:

```{r}

tibble(
    Feature = c("crimes.Buffer (600 ft)", "crimes_500ft", "crime_nn3", "dist_downtown_mi"),
    Type = c("Buffer Count", "Buffer Count", "kNN Distance", "Point Distance"),
    "what It Measures" = c("Number of Crimes Near House", "Crimes Within 500 ft", "Avg. Distance to 3 Nearest Crimes", "Miles From Downtown Boston")
)

```

Model Comparison: Adding Spatial Features

```{r}

boston.sf <- boston.sf %>%
  mutate(Age = 2015 - YR_BUILT)  

# Model 1: Structural only
model_structural <- lm(SalePrice ~ LivingArea + R_BDRMS + Age, 
                       data = boston.sf)

# Model 2: Add spatial features
model_spatial <- lm(SalePrice ~ LivingArea + R_BDRMS + Age +
                    crimes_500ft + crime_nn3 + dist_downtown_mi,
                    data = boston.sf)

# Compare
cat("Structural R²:", round(summary(model_structural)$r.squared, 4),
    "\nWith spatial R²:", round(summary(model_spatial)$r.squared, 4),
    "\nImprovement:", round(summary(model_spatial)$r.squared - summary(model_structural)$r.squared, 4)
)

```

PART 3: FIXED EFFECTS

What Are Fixed Effects?
* Fixed Effects = Categorical variables that capture all unmeasured characteristics of a group, often called “unobservables”

In hedonic models:
* Each neighborhood gets its own dummy variable
* Captures everything unique about that neighborhood we didn’t explicitly measure
* We technically already did this when I went over categorical data!

What They Capture:
* School quality
* “Prestige” or reputation
* Walkability
* Access to jobs
* Cultural amenities
* Things we can’t easily measure

Code:

```{r}

# Add neighborhood fixed effects
reg5 <- lm(
  SalePrice ~ LivingArea + Age + 
              crimes_500ft + 
              #parks_nn3 + 
              as.factor(name),  # FE
  data = boston.sf
)

summary(reg5)

```

R adds a dummy variable for each neighborhood automatically
* Behind the scenes, R creates dummies:
* is_BackBay = 1 if Back Bay, 0 otherwise
* is_Beacon = 1 if Beacon Hill, 0 otherwise
* is_Allston = 1 if Allston, 0 otherwise
* ... (R drops one as reference category)

Interpretation Example:

Coefficients:
(Intercept)           50000   ← $50000 is the reference price 
LivingArea              150
nameBack_Bay          85000   ← $85k premium vs. reference (i.e. if you're in Back Bay you're this much more than ref. 'hood)
nameBeacon_Hill      125000   ← $125k premium  
nameAllston          -15000   ← $15k discount

Each coefficient = price premium/discount for that neighborhood (holding all else constant)

Why Use Fixed Effects?

1. Dramatically Improve Prediction

  Model Comparison (R²):
  - Structural only:     0.58
  - + Spatial features:  0.67
  - + Fixed Effects:     0.81 ✓

  Why such a big jump?
    * Neighborhoods bundle many unmeasured factors
      * School districts
      * Job access
      * Amenities
      * “Cool factor”

2. Coefficients Change
  Crime coefficient:
    Without FE: -$125/crime
    With FE: -$85/crime
  
  Why?
    Without FE: captured neighborhood confounders too
    With FE: neighborhoods “absorb” other differences
    Now just the crime effect

TRADEOFF:
  * FE are powerful but they’re a black box - we don’t know WHY Back Bay commands a premium
  
Comparison of all models:

```{r}

# Model 3: Structural Only
reg3 <- lm(SalePrice ~ LivingArea + Age + R_FULL_BTH, 
           data = boston.sf)

# Model 4: Add Spatial Features  
reg4 <- lm(SalePrice ~ LivingArea + Age + R_FULL_BTH +
                       crimes_500ft + crime_nn3+ dist_downtown_mi,
           data = boston.sf)


# Model 5: Add Fixed Effects
reg5 <- lm(SalePrice ~ LivingArea + Age + R_FULL_BTH +
                       crimes_500ft + crime_nn3+ dist_downtown_mi +
                       as.factor(name),
           data = boston.sf)


library(modelsummary)

#compare model fit
modelsummary(
  list("Structural" = reg3, 
       "Spatial Features" = reg4, 
       "Fixed Effects" = reg5),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)

```

But in-sample R² can be misleading! We need cross-validation…

PART 5: CROSS VALIDATION (WITH CATEGORICAL VARIABLES)

Three common validation approaches: (Recap from last week)
  1. Train/Test Split - 80/20 split, simple but unstable
  2. k-Fold Cross-Validation - Split into k folds, train on k-1, test on 1, repeat
  3. LOOCV - Leave one observation out at a time (special case of k-fold)

Here we'll use k-fold CV to compare out hedonic models.

Why?
  * Tells us how well model predicts NEW data
  * More honest than in-sample R^2
  * Helps detect overfitting

```{r}

library(caret)
ctrl <- trainControl(method = "cv", number = 10)

# Model 1: Structural
cv_m1 <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH,
  data = boston.sf, method = "lm", trControl = ctrl
)

# Model 2: + Spatial
cv_m2 <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH + crimes_500ft + crime_nn3,
  data = boston.sf, method = "lm", trControl = ctrl
)

# Model 3: + Fixed Effects (BUT WAIT - there's a (potential) problem!)
cv_m3 <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH + crimes_500ft + crime_nn3 + 
              as.factor(name),
  data = boston.sf, method = "lm", trControl = ctrl
)

# Compare
tibble(
  Model = c("Structural", "Spatial", "Fixed Effects"),
  RMSE = c(cv_m1$results$RMSE, cv_m2$results$RMSE, cv_m3$results$RMSE)
)

```

THE PROBLEM: Sparse Categories

When CV Fails with Categorical Variables:
  * You might see this error: Error in model.frame.default: factor ‘name’ has new level ‘Fenway’

What happened?

  * Random split created 10 folds
  * All “West End” sales ended up in ONE fold (the test fold)
  * Training folds never saw “Fenway”
  * Model can’t predict for a category it never learned
  * When neighborhoods have very few sales (<10), random CV splits can put all instances in the same fold, breaking the model.
  
ALWAYS check the data first before CV with categorical variables:

```{r}

# ALWAYS run this before CV with categorical variables
category_check <- boston.sf %>% 
  st_drop_geometry() %>%
  count(name) %>%
  arrange(n)

print(category_check)

```

Typical output:

name                n
Bay Village         1  ⚠️ Problem!
Allston             6  ⚠️ Risky
Mission Hill       14  ⚠️ Borderline
South Boston       78  ✓ Safe
Dorchester        340  ✓ Safe

Rule of thumb: Categories with n < 10 will likely cause CV problems

SOLUTION: Group Small Neighborhoods
  * Tradeoff: You lose granularity for small neighborhoods, but avoid CV crashes.

```{r}

# Step 1: Add count column
boston.sf <- boston.sf %>%
  add_count(name)

# Step 2: Group small neighborhoods
boston.sf <- boston.sf %>%
  mutate(
    name_cv = if_else(
      n < 10,                       # If fewer than 10 sales
      "Small_Neighborhoods",        # Group them
      as.character(name)            # Keep original
    ),
    name_cv = as.factor(name_cv)
  )

# Step 3: Use grouped version in CV
cv_model_fe <- train(
  SalePrice ~ LivingArea + Age + crimes_500ft + 
              as.factor(name_cv),   # Use name_cv, not name!
  data = boston.sf,
  method = "lm",
  trControl = trainControl(method = "cv", number = 10)
)

```


ALTERNATIVE SOLUTION: Drop Sparce Categories
(For when grouping doesn't make sense)

  * Consider carefully: Which neighborhoods are you excluding? Often those with less data are marginalized communities.
  * Document what you removed and why.

```{r}

# Remove neighborhoods with < 10 sales
neighborhood_counts <- boston.sf %>% 
  st_drop_geometry() %>%
  count(name)

keep_neighborhoods <- boston.sf %>%
  filter(n >= 10) %>%
  pull(name)

boston_filtered <- boston.sf %>%
  filter(name %in% keep_neighborhoods)

cat("Removed", nrow(boston.sf) - nrow(boston_filtered), "observations")

```

Prof's recommended workflow:

```{r}

# 1. Check category sizes
boston.sf %>%
  st_drop_geometry() %>%
  count(name) %>%
  arrange(n) %>%
  print()

# 2. Group if needed
boston.sf <- boston.sf %>%
  add_count(name) %>%
  mutate(
    name_cv = if_else(n < 10, "Small_Neighborhoods", as.character(name)),
    name_cv = as.factor(name_cv)
  )

# 3. Set up CV
ctrl <- trainControl(method = "cv", number = 10)

# 4. Use grouped neighborhoods in ALL models with FE
model <- train(
  SalePrice ~ LivingArea + Age + crimes_500ft + as.factor(name_cv),
  data = boston.sf,
  method = "lm",
  trControl = ctrl
)

# 5. Report
cat("\n\n\n10-fold CV RMSE:", round(model$results$RMSE, 0), "\n")

```

Full Model Comparison With CV:

```{r}

library(caret)

# Prep data
boston.sf <- boston.sf %>%
  add_count(name) %>%
  mutate(name_cv = if_else(n < 10, "Small_Neighborhoods", as.character(name)),
         name_cv = as.factor(name_cv))

ctrl <- trainControl(method = "cv", number = 10)

# Compare models
cv_structural <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH,
  data = boston.sf, method = "lm", trControl = ctrl
)

cv_spatial <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH + crimes_500ft + crime_nn3,
  data = boston.sf, method = "lm", trControl = ctrl
)

cv_fixedeffects <- train(
  SalePrice ~ LivingArea + Age + R_FULL_BTH + crimes_500ft + crime_nn3 + 
              as.factor(name_cv),
  data = boston.sf, method = "lm", trControl = ctrl
)

# Results
data.frame(
  Model = c("Structural", "+ Spatial", "+ Fixed Effects"),
  RMSE = c(cv_structural$results$RMSE, 
           cv_spatial$results$RMSE, 
           cv_fixedeffects$results$RMSE)
)

```

EXPECTED RESULTS:

Typical pattern:

Model              RMSE      Interpretation
Structural        $533,330.60   Baseline - just house characteristics
+ Spatial         $500,421.5    Adding location features helps!
+ Fixed Effects   $347,261.30   Neighborhoods capture a LOT

Note:
  * These values are kind of ginormous - remember RMSE squares big errors, so outliers can have a really large impact

Key Insight: 
  * Each layer improves out-of-sample prediction, with fixed effects providing the biggest boost

Why? 
  * Neighborhoods bundle many unmeasured factors (schools, amenities, prestige) that we can’t easily quantify individually


Investigating Errors:

```{r}

ggplot(boston.sf, aes(x = SalePrice)) +
  geom_histogram(bins = 50) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Distribution of Sale Prices")
summary(boston.sf$SalePrice)

boston.sf %>%
  st_drop_geometry() %>%
  arrange(desc(SalePrice)) %>%
  select(SalePrice, LivingArea, name) %>%
  head(10)


ggplot(boston.sf, aes(y = SalePrice)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Sale Price Outliers")

```

**Look for:**
- Prices over $2-3 million (could be luxury condos or errors)
- Prices near $0 (data errors)
- Long right tail in histogram

What to Do?
  * Log-transform the skewed dependent variable
  
  * Interpreting Log Models
  
    * RMSE is now in log-dollars (hard to interpret)
    * To convert back: exp(predictions) gives actual dollars
    * Coefficients now represent PERCENTAGE CHANGES, not dollar changes
    * This is standard practice in hedonic modeling!
    
    
Tips For Success In Model Building:

DO: 

  1. Start simple, add complexity
  2. Check for NAs with sum(is.na())
  3. Test on small subset first
  4. Comment your code
  5. Check coefficient signs
  6. Use glimpse() and summary()
  
DON'T:

  1. Add 50 variables at once
  2. Ignore errors
  3. Forget st_drop_geometry() for non-spatial operations
  4. Skip sparse category check

COMMON ERRORS:

  1. “Factor has new levels” 
      → Group sparse categories
  
  2. “Computationally singular” 
      → Remove collinear variables
  
  3. Very high RMSE 
      → Check outliers, scale
  
  4. CV takes forever 
      → Simplify model or reduce folds
  
  5. Negative R² 
      → Model worse than mean, rethink variables
